A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-12 evaluation.tex
more

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances of Parikh
automata intersection problems generated by the \OstrichPlus{} string constraint
solver when solving the PyEx benchmarks~\cite{pyex}. After generating an
initial~\InitialNrBenchmarks{}, we remove~\NrTrivial{} instances solved in under
five seconds by baseline as well as~\NrInvalid{} instances that produced errors
during experimentation. We also attempted to benchmark instances generated by
\Ostrich{} solving the kaluza-len (\numprint{38227}~instances) and pyex-len
(\numprint{791}~instances) benchmark suites but discarded them since they were
all trivial. For instance, on a laptop, \Calculus{} could solve every instance
from kaluza-len in under five seconds (mean about 0.1s). The benchmarks are run
on commit~\texttt{\commit}.

@@

\begin{table}
  \centering
  \input{graphs/solved_pivot_table.tex}
  \caption{Number of successful results within a timeout of \RuntimeTimeout{}.
  Instances solved by no backend within the timeout (about half of the set) are
  omitted from the table.}\label{tab:solve-status}
\end{table}

>>>>

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances of Parikh
automata intersection problems generated by the \OstrichPlus{} string constraint
solver~\cite{ostrich-plus} when solving the PyEx benchmarks, 
which are string constraints from symbolic execution that are known
to be hard for many solvers~\cite{pyex}. After generating an
initial~\InitialNrBenchmarks{}, we remove~\NrTrivial{} instances solved in under
five seconds by baseline as well \Fudge{as~\NrInvalid{} instances that produced errors
during experimentation}.
% Better explanation needed why we removed the benchmarks
 We also attempted to benchmark instances generated by
\OstrichPlus{} solving the Kaluza benchmarks~\cite{Saxena10:kaluza}
(\numprint{38227}~instances),
\iffalse
and pyex-len
(\numprint{791}~instances) benchmark suites\fi
but discarded them since they all turned out to be trivial for our tool
\Catra{}: every instance
could be solved in under five seconds, with a mean runtime of about 0.1s.
\Fudge{The benchmarks are run
on commit~\texttt{\commit}.}
% Commit of what?

@@

\begin{table}
  \begin{center}
  \input{graphs/solved_pivot_table.tex}
  \end{center}
  \caption{Number of successful results within a timeout of \RuntimeTimeout{}.
  Instances solved by no backend within the timeout (about half of the set) are
  omitted from the table.}\label{tab:solve-status}
\end{table}

### 2023-07-12 extensions.tex
Evaluation + kill your darlings
more

\Calculus{} can be accelerated for some instances by adding rules for
backjumping. In particular, key connectedness constraints for an automaton can
be learnt. In that case, when discovering that a transition $\Transition$ of an
automaton $\Automaton$ under a certain transition variable $\Filter$ has become
unreachable, we can learn the clause $\Connected(\Automaton, \Filter) \land
\sum_{\Transition' \in C} \Filter(\Transition') = 0 \implies \Transition = 0$ where
$\Dominates(C, \Automaton, \Transition)$ for some minimal $C$. Note that this
captures both forward (unrachable from the initial state) and
backward (does not reach an accepting state) unreachability. Bidirectional cut
learning is implemented in \Catra, and has provided a slight improvement in
performance on some instances.

The other source of clauses to learn is the \Materialise{} rule. Whenever an
attempt to materialise a product of two ($\Filter$-filtered) automata
$\Automaton_1 \times \Automaton_2$ produces an empty product, we can determine
the cause of the failure with respect to the automata and their respective
transition variables, and learn no-good combinations that can never be part of a
model. In order to be able to do this, we need a few semantic predicates to
record the state of the calculation, as well as a system of disambiguating
automata, since it is possible to arrive at the same automaton by multiple
combinations of decisions and products. In \Catra{}, we use a number of
additional predicates to record the status of the materialisation of products,
to separate instances of our main predicate, and to register the mapping between
automata and their respective transition $\Filter$ terms, described in
\cref{sec:implementation}. However, at this point only rudimentary no-good
learning is implemented.

@@

\subsection{Transducers}\label{sec:ext:transducers}

@@

Implementing such a calculus is straightforward in \Calculus, since the
definition of products was left out of the definition. All that would be needed
is an appropriate update of the definitions of products. Simiarly, \Catra was
written with modularity in mind, and it should be straightforward to extend both
the input grammar and automata implementations to accommodate multi-track
automata with arbitrary synchronisation.

@@

    A preliminary implementation of this approach is available as part
    of~\Catra, but has been neither optimised nor fully tested.

>>>>

\Calculus{} can be accelerated for some instances by adding rules for
backjumping. In particular, key connectedness constraints for an automaton can
be learnt. In that case, when discovering that a transition $\Transition$ of an
automaton $\Automaton$ under a certain transition variable $\Filter$ has become
unreachable, we can learn the clause $\Connected(\Automaton, \Filter) \land
\bigwedge_{\Transition' \in C} \Filter(\Transition') = 0 \implies \Transition = 0$ where
$\Dominates(C, \Automaton, \Transition)$ for some minimal $C$. Note that this
captures both forward (unrachable from the initial state) and
backward (does not reach an accepting state) unreachability. Bidirectional cut
learning is implemented in \Catra, and has provided a slight improvement in
performance on some instances.

The other source of clauses to learn is the \Materialise{} rule. Whenever an
attempt to materialise a product of two ($\Filter$-filtered) automata
produces an empty product, we can determine
the cause of the failure with respect to the automata and their respective
transition variables, and learn no-good combinations that can never be part of a
model. In order to be able to do this, we need a few semantic predicates to
record the state of the calculation, as well as a system of disambiguating
automata, since it is possible to arrive at the same automaton by multiple
combinations of decisions and products. In \Catra{}, we use a number of
additional predicates to record the status of the materialisation of products,
to separate instances of our main predicate, and to register the mapping between
automata and their respective transition $\Filter$ terms, described in
\cref{sec:implementation}. However, at this point only rudimentary no-good
learning is implemented.

@@

\iffalse
\subsection{Transducers}\label{sec:ext:transducers}

@@

Implementing such a calculus is straightforward in \Calculus, since the
definition of products was left out of the definition. All that would be needed
is an appropriate update of the definitions of products. Similarly, \Catra was
written with modularity in mind, and it should be straightforward to extend both
the input grammar and automata implementations to accommodate multi-track
automata with arbitrary synchronisation.

@@

    A preliminary implementation of this approach is available as part
    of~\Catra, but has been neither optimised nor fully tested.
\fi

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2023-07-12 implementation.tex
more

In satisfaction mode, supported by all backends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting \Sat{} with register
assignments or \Unsat{} much like traditional~SAT- or SMT solvers would.
Additionally, baseline and \Calculus{} also support generating the Presburger
formula describing the constraints of the input file. Baseline uses standard
quantifier elimination, and \Calculus{} uses the method described in
\cref{sec:finding-the-image}.

Since \Princess{} does not support the multiple-arity predicates used in
\Calculus{}, we have implemented variable-length arguments using additional
helper predicates. These are $\Unused{}(\Automaton)$, which marks an automaton
as unused in any product, and $\TransitionMask{}(\Automaton, \Transition,
\Filter(\Transition))$ which associates a transition $\Transition$ and automaton
$\Automaton$ with its corresponding transition variable. Additionally, we
associate each predicate with an instance variable to tie helper predicates
to the $\Image$ predicate they originated from.

@@

As an optimisation, our automata (including intermittent products) have dead
states eliminated during construction. Any automaton we produce contains only
states that are both reachable from the initial state and has a path to an
accepting state. We never perform any other minimisation on the automata for
either backend. More complex minimisation was left out since performing
minimisation on automata with counters is non-trivial.

@@

Product materialisation is the final piece of the puzzle. In the current
implementation we put off computing intermediate products until six transition
variables of one of the automata is known to be either present or absent. The
number was chosen experimentally. The other automaton for the product is
selected randomly.

>>>>

In satisfaction mode, supported by all backends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting \Sat{} with register
assignments or \Unsat{} much like traditional~SAT- or SMT solvers would.
Additionally, baseline and \Calculus{} also support generating the Presburger
formula describing the constraints of the input file, i.e., computing
a closed-from representation of the complete Parikh image.
% Baseline uses standard
%quantifier elimination, and \Calculus{} uses the method described in
%\cref{sec:finding-the-image}.

Since \Princess{} does not support the multiple-arity predicates used
in \Calculus{}, we have implemented variable-length arguments using
additional helper predicates. These are $\Unused{}(\Automaton)$, which
marks an automaton as unused in any product, and
$\TransitionMask{}(\Automaton, \Transition, \Filter(\Transition))$
which associates a transition $\Transition$ and automaton $\Automaton$
with its corresponding transition
variable~$\Filter(\Transition)$. Additionally, we associate each
predicate with an instance variable to tie helper predicates to the
$\Image$ predicate they originated from.

@@

As an optimisation, our automata (including intermittent products) have dead
states eliminated during construction. Any automaton we produce contains only
states that are both reachable from the initial state and have a path to an
accepting state. We never perform any other minimisation on the automata for
either backend. More complex minimisation was left out since performing
minimisation on automata with counters is non-trivial.

@@

Product materialisation is the final piece of the puzzle. In the current
implementation we put off computing intermediate products until six transition
variables of one of the automata is known to be either \Fudge{present or absent}. The
number was chosen experimentally. The other automaton for the product is
selected randomly.

% this is unclear; don't we delay computing products until one of the
% automata is (almost) completely defined?
